Spatial Generalized Linear Models in \proglang{R} Using \pkg{spmodel}

### 2025-11-26 inst/manuscript/manuscript.tex
some updates

\Abstract{
Non-Gaussian data are common in practice and include binary, count,
skewed, and proprtion data types. Often, non-Gaussian data are modeled
using a generalized linear model (GLM). GLMs typically assume that
observations are independent of one another. This is an impractical
assumption for spatial data, as nearby observations tend to be more
similar than distant ones. The \pkg{spmodel} package in \proglang{R}
provides a suite of tools for fitting spatial generalized linear models
(SPGLMs) to non-Gaussian data and making spatial predictions (i.e.,
Kriging). SPGLMs for point-referenced data (x- and y-coordinates) are
fit using the \code{spglm()} function, while SPGLMs for areal (lattice,
polygon) data are fit using the \code{spgautor()} function. Both
\code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood
which marginalizes over the model's fixed effects and latent mean while
formally incorporating spatial covariance. The inputs and outputs of
\code{spglm()} and \code{spgautor()} closely resemble the \code{glm()}
function from base \proglang{R}, easing the transition from GLMs to
SPGLMs. \pkg{spmodel} provides and builds upon several commonly used
helper functions for model building like \code{summary()},
\code{plot()}, \code{fitted()}, and \code{tidy()}, among others. Spatial
predictions of the latent mean at unobserved locations are obtained
using \code{predict()} or \code{augment()}. \pkg{spmodel} accommodates
myriad advanced modeling features like geometric anisotrpoy, nonspatial
random effects, analysis of variance, and more. Throughout, we use
\pkg{spmodel} to fit SPGLMs to moose presence and counts in Alaska,
United States (US), skewed conductivity data in the Southwestern US,
harbor seal abundance trends in Alaska, US, and voter turnout rates in
Texas, US.
}

\Keywords{autoregressive model, geostatistical model, inverse link
function, Poisson regression, link function, logistic
regression, spatial covariance, spatial correlation}
\Plainkeywords{autoregressive model, geostatistical model, inverse link
function, Poisson regression, link function, logistic
regression, spatial covariance, spatial correlation}

@@

In practice, non-Gaussian data (e.g., binary, count, skewed, and
proportion data) are ubiquitous. Non-Gaussian data that belong to an
exponential family can be naturally modeled using a generalized linear
model (GLM) regression framework
\citep{nelder1972generalized, mccullagh1989generalized}. In a GLM, an
\(n \times 1\) response variable \(\mathbf{y}\) belongs to a statistical
distribution (e.g., Binomial, Poisson) with some mean and variance.
Often, the analysis goal is to study the impact of a linear function of
several explanatory variables on the mean of \(\mathbf{y}\) through a
GLM. In this context, the latent (i.e., unobserved) mean of
\(\mathbf{y}\), \(\boldsymbol{\mu}\), is linked to these explanatory
variables via a link function: \begin{align}\label{eq-glm}
f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
\end{align} where for a sample size \(n\), \(f(\cdot)\) is a link
function that connects \(\boldsymbol{\mu}\) to \(\mathbf{w}\),
\(\mathbf{X}\) is the \(n \times p\) design matrix of explanatory
variables, and \(\boldsymbol{\beta}\) is the \(p \times 1\) vector of
fixed effects. While the mean is typically constrained in some way
(e.g., if a probability, between zero and one), the link function
generally makes \(\mathbf{w}\) unconstrained. Common link functions
include the log odds (i.e., logit) link for binary and proportion data
and the log link for count and skewed data. Equation\(~\)\ref{eq-glm}
can also be written in terms of the inverse link function,
\(f^{-1}(\cdot)\): \begin{align}\label{eq-glm2}
\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta} \equiv f^{-1}(\mathbf{w}) = f^{-1}(\mathbf{X} \boldsymbol{\beta}).
\end{align}

The GLM fixed effects (\(\boldsymbol{\beta}\)) are typically estimated
via maximum likelihood \citep{chambers1992S}. It is often convenient to
compute the maximum likelihood estimates using the iteratively
reweighted least squares (IRWLS) algorithm \citep{wood2017generalized},
which is the approach used by the \code{glm()} function in the
\proglang{R} programming language \citep{rcore2024}. GLMs add an
additional layer of complexity compared to linear regression models, as
the left-hand side of Equation\(~\)\ref{eq-glm} is a function of the
mean of \(\mathbf{y}\) rather than \(\mathbf{y}\) itself (as in linear
regression models).

The standard GLM assumes the elements of \(\mathbf{y}\) are independent.
This independence assumption is typically impractical for spatial data.
For spatial data, nearby observations tend to be more similar than
distant observations \citep{tobler1970computer}, which leads to positive
spatial covariance. The consequences of ignoring spatial covariance in
statistical models for spatial data can be severe and include imprecise
parameter estimates as well as misleading standard errors that inflate
Type-I error rates and decrease power \citep{zimmerman2024spatial}.

An approach for handling spatial data using a GLM is to assume the
elements of \(\mathbf{w}\) exhibit covariance that varies spatially and
nonspatially. This is achieved by adding to Equation\(~\)\ref{eq-glm}
two random effects, \(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\).
The random effect \(\boldsymbol{\tau}\) is an \(n \times 1\) column
vector of spatially dependent random errors. We assume that
\(\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}\) and
\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}\), where
\(\text{E}(\cdot)\) and \(\text{Cov}(\cdot)\) denote expectation and
covariance, respectively. The variance parameter \(\sigma^2_\tau\)
controls the magnitude of spatial covariance and is often called a
partial sill. The matrix \(\mathbf{R}\) is an \(n \times n\) spatial
correlation matrix that depends on a range parameter controlling the
distance-decay rate of the spatial correlation. One example of a spatial
covariance matrix is the ``exponential,'' which is given by
\begin{align}\label{eq-spcov-exp}
  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{\tau} \mathbf{R}_{exp} =  \sigma^2_{\tau} \exp(-\mathbf{H}/\phi),
\end{align} where \(\mathbf{H}\) is a matrix of pairwise distances among
the elements of \(\mathbf{y}\) and \(\phi\) is the range parameter. From
Equation\(~\)\ref{eq-spcov-exp}, as the distance between two elements of
\(\mathbf{y}\) increases, the spatial covariance decreases, which
reflects intuition. Moreover, as the range parameter, \(\phi\),
increases, the strength of spatial dependence increases
(Figure\(~\)\ref{fig-range}). The random effect
\(\boldsymbol{\epsilon}\) is an \(n \times 1\) column vector of
independent random errors. We assume that
\(\text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}\) and
\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\epsilon \mathbf{I}\), where
\(\mathbf{I}\) is an \(n \times n\) identity matrix. The variance
parameter \(\sigma^2_\epsilon\) controls the magnitude of nonspatial
variability (i.e., fine-scale variation) and is often called a nugget.
Often in spatial statistics, quantities are explicitly referenced with
respect to \(\mathbf{s}\), a vector of spatial coordinates indexing the
observation \citep{cressie1993statistics}. For example, \(\mathbf{y}\)
and \(\mathbf{X}\) may instead be written \(\mathbf{y}(\mathbf{s})\) and
\(\mathbf{X}(\mathbf{s})\), respectively. We acknowledge the utility of
this nomenclature but drop the explicit dependence on \(\mathbf{s}\) for
simplicity of notation moving forward.

@@

Fitting and using SPGLMs is challenging both conceptually and
computationally \citep{bolker2009generalized}. Recently, however, there
have been numerous, significant advances in \proglang{R} software that
have made these models more accessible to practitioners. The \pkg{brms}
\citep{burkner2017brms}, \pkg{carBayes} \citep{lee2013carbayes},
\pkg{ngspatial} \citep{hughes2020ngspatial}, \pkg{R-INLA}
\citep{lindgren2015bayesian} and \pkg{inlabru} \citep{bachl2019inlabru},
\pkg{spBayes} \citep{finley2007spbayes}, \pkg{spOccupancy}
\citep{doser2022spoccupancy}, \pkg{spAbundance}
\citep{doser2024spabundance}, and \pkg{spNNGP} \citep{finley2002spnngp}
packages take a Bayesian approach, either directly sampling from
posterior distributions of parameters (e.g., using MCMC) or
approximating them. A benefit of Bayesian approaches is that prior
information can be incorporated and uncertainty quantification of
parameter estimates is straightforward. However, Bayesian approaches,
especially those using MCMC, can be computationally expensive. In order
to reduce computation time, many of these packages work with the
precision matrix instead of the covariance matrix so that
computationally expensive matrix inversion is not required. For example,
\pkg{R-INLA} uses the precision matrix and tends to be very fast.
Working with precision matrices, however, can be more restrictive and
less intuitive than working directly with the covariance matrix. The
\pkg{FRK} \citep{sainsbury2024modeling}, \pkg{glmmTMB}
\citep{brooks2017glmmtmb}, \pkg{hglm} \citep{ronnegard2010hglm},
\pkg{mgcv} \citep{wood2017generalized}, and \pkg{spaMM}
\citep{rousset2014spamm} packages directly use Laplace,
quasi-likelihood, or reduced-rank approaches to estimate parameters.
These direct approaches tend to be computationally efficient, as they
don't rely on MCMC sampling. In contrast to the Bayesian approach, a
drawback of these direct approaches is that prior information cannot be
formally incorporated and covariance parameter uncertainty is more
challenging to quantify. The \pkg{sdmTMB} \citep{anderson2024sdmtmb}
package combines elements of \pkg{R-INLA}, \pkg{glmmTMB}, and Gaussian
Markov random fields to fit a wide variety of SPGLMs, while
\pkg{tinyVAST} \citep{thorson2025tinyVAST} extends some of these models
to multivariate or (dynamic) structural equation models.

Building from \citet{evangelou2011estimation} and
\citet{bonat2016practical}, \citet{ver2024marginal} proposed a novel
approach for fitting SPGLMs that leverages the Laplace approximation
while marginalizing over both the latent \(\mathbf{w}\) and the fixed
effects (\(\boldsymbol{\beta}\)) and accommodating general covariance
structures, including spatial ones. This approach performed efficiently
in a variety of simulation settings, generally having appropriate
confidence interval coverage for the fixed effects and prediction
interval coverage for \(\mathbf{w}\) at new locations. The approach
performed similarly to the Bayesian SPGLM approach in \pkg{spBayes} and
the automatic differentiation SPGLM approach in \pkg{glmmTMB} but was
much faster. At small sample sizes, the approach outperformed the
approximate Bayesian SPGLM approach in \pkg{R-INLA} and had similar
computational times. For moderate sample sizes, it performed similarly
to \pkg{R-INLA}, though \pkg{R-INLA} was faster. The novel Laplace
approach is particularly attractive for two reasons. First, it is
general enough that it can be applied to any covariance structure (not
just spatial). Second, after estimating the covariance parameters,
analytical solutions exist for the fixed effects (and their standard
errors) as well as predictions of the latent \(\mathbf{w}\) at new
locations (and their standard errors).

The \pkg{spmodel} \proglang{R} package \citep{dumelle2023spmodel}
recently made public a full set of modeling tools for SPGLMs fit using
the novel Laplace approach described by \citet{ver2024marginal}. These
modeling tools are approachable and mirror the familiar \code{glm()}
syntax from base-\proglang{R}, making the transition from GLMs to SPGLMs
relatively seamless. The \code{spglm()} function fits SPGLMs for
point-referenced data (e.g., x- and y-coordinates representing point
locations in a field; these models are sometimes called
``geostatistical'' models), while the \code{spgautor()} function fits
SPGLMs for areal data (e.g., polygon boundaries representing geographic
subsets of a region; these models are sometimes called
``autoregressive'' models). For both point-referenced data and areal
supports, \pkg{spmodel} supports the binomial distribution for binary
data, Poisson and negative binomial distributions for count data, Gamma
and inverse Gaussian distributions for skewed data, and the beta
distribution for proportion data. There are 20 different spatial
covariance structures available including the exponential, Gaussian, and
spherical for point-referenced data (Figure\(~\)\ref{fig-type}) and the
conditional autoregressive, and simultaneous autoregressive structures
for areal data. \pkg{spmodel} provides tools for commonly used model
summaries, visualizations, and diagnostics (e.g., Cook's distance) using
standard \proglang{R} helper functions like \code{summary()},
\code{plot()}, \code{fitted()}, and \code{tidy()}, among others.
\pkg{spmodel} also provides tools to predict \(\mathbf{w}\) at new
locations and quantify uncertainty in those prediction using
\code{predict()} and \code{augment()}. This core functionality, combined
with several advanced features we describe throughout the manuscript,
enables \pkg{spmodel} to introduce novel, important SPGLM modeling tools
previously missing from the existing \proglang{R} ecosystem.

@@

Of the existing \proglang{R} packages for SPGLMs, \pkg{spmodel} (version
0.11.0) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in
terms of scope and feel. Both packages use similar syntax as
\code{glm()}, accommodate flexible \code{formula} arguments (e.g.,
offsets, splines), handle spatial covariance that decays at different
rates in different directions (i.e., geometric anisotropy), incorporate
nonspatial random effects, support other \proglang{R} packages for
modeling like \pkg{broom} \citep{robinson2021broom, kuhn2022tidy},
\pkg{emmeans} \citep{lenth2024emmeans}, and \pkg{car}
\citep{fox2019car}, and have tools for model summaries, prediction, and
simulating data. There are some notable differences between the two
packages, however. \pkg{sdmTMB} supports several additional GLM
distributions like the Tweedie, supports Hurdle models, and can
incorporate prior information through Bayesian applications.
\pkg{sdmTMB} also provides tools for working with temporal data and
spatiotemporal data and provides enhanced visualizations of the model's
marginal effects. \pkg{sdmTMB} does require a preprocessing step of
constructing a mesh prior to model fitting (using the stochastic partial
differential equation approach), and the density of the mesh can affect
model results and computational complexity. On the other hand,
\pkg{spmodel} does not require the construction of a mesh prior to model
fitting. \pkg{spmodel} supports 20 different spatial covariances and
models them directly, rather than using a precision matrix approximation
to the Matérn spatial covariance as in \pkg{sdmTMB}. \pkg{spmodel} can
model data directly using neighborhood distance and autoregressive
models, rather than relying on the polygon centroid (as in
\pkg{sdmTMB}), which may not be within the polygon's boundaries.
\pkg{spmodel} provides experimental design tools (e.g., analysis of
variance, contrasts), supports \pkg{sf} objects in modeling and
prediction functions \citep{pebesma2018sf}, has several specialized
model diagnostics like leverage values and Cook's distances, and has
analytic solutions for fixed effect and prediction standard errors.
Other similarities and differences do exist between \pkg{sdmTMB} and
\pkg{spmodel}, and both packages continue to evolve. Overall, we believe
that these packages are complementary and enhance the suite of SPGLM
tools accessible to practitioners.

The rest of this article is organized as follows. In
Section\(~\)\ref{sec-spglm}, we provide some background for the SPGLM
fitting and prediction routines in \pkg{spmodel}. In
Section\(~\)\ref{sec-applications}, we provide an overview of core SPGLM
functionality in \pkg{spmodel} by modeling moose presence in Alaska,
United States (US). In Section\(~\)\ref{sec-applications2}, we model
moose counts in Alaska, US; skewed lake conductivity in the Southwestern
US; harbor seal abundance trend behavior in Alaska, US; and voter
turnout rates in Texas, US. And in Section\(~\)\ref{sec-discussion}, we
end with a discussion synthesizing \pkg{spmodel}'s contributions to the
analysis of SPGLMs in \proglang{R}.

@@

The novel Laplace approach implemented in \pkg{spmodel} formally
maximizes a hierarchical GLM likelihood
\citep{lee1996hierarchical, wood2017generalized}, making
likelihood-based statistics for model comparison like AIC
\citep{akaike1974new}, AICc \citep{hoeting2006model}, BIC
\citep{schwarz1978estimating}, deviance
\citep{mccullagh1989generalized}, and likelihood ratio tests available.
These types of statistics are not available for quasi-likelihood
\citep{wedderburn1974quasi, breslow1993approximate} or pseudo-likelihood
approaches \citep{wolfinger1993generalized}, which only specify the
first two moments of a distribution. Next, we describe a brief overview
of the approach and how it can be used for several primary data analysis
tasks \citep{tredennick2021practical} like model comparison, parameter
estimation, inference, model diagnostics, and prediction. Then in
Section\(~\)\ref{sec-applications} and
Section\(~\)\ref{sec-applications2}, we show how to use \pkg{spmodel} to
carry out these primary data analysis tasks with various data sets.

@@

  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{w}^{*}), \varphi] [\mathbf{w}^{*} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{w}^{*}}|^{-1/2}.
\end{align}

Maximizing the natural logarithm of Equation\(~\)\ref{eq-marginal6}
requires a doubly iterative process over 1) \(\boldsymbol{\theta}\) and
\(\varphi\) and 2) the latent \(\mathbf{w}\) (to find each set of
\(\mathbf{w}^{*}\)). This doubly iterative process eventually yields the
marginal restricted maximum likelihood estimators \(\hat{\varphi}\) and
\(\hat{\boldsymbol{\theta}}\) and their respective values of
\(\mathbf{w}^{*}\), which we call \(\hat{\mathbf{w}}\). Maximizing the
marginal restricted likelihood requires repeatedly evaluating
\(\boldsymbol{\Sigma}^{-1}\), \(\mathbf{g}\), and \(\mathbf{G}\); see
\citet{ver2024marginal} for more details and comparisons to other
approaches as well as forms of \(\mathbf{g}\) and \(\mathbf{G}\) for
various response distributions.

\subsection{Estimating fixed effects}\label{estimating-fixed-effects}

>>>>

\Abstract{
Non-Gaussian data are common in practice and include binary, count,
skewed, and proportion data types. Often, non-Gaussian data are modeled
using a generalized linear model (GLM). GLMs typically assume that
observations are independent of one another. This is an impractical
assumption for spatial data, as nearby observations tend to be more
similar than distant ones. The \pkg{spmodel} package in \proglang{R}
provides a suite of tools for fitting spatial generalized linear models
(SPGLMs) to non-Gaussian data and making spatial predictions (i.e.,
Kriging). SPGLMs for point-referenced (x- and y-coordinates) support are
fit using the \code{spglm()} function, while SPGLMs for areal (lattice,
polygon) support are fit using the \code{spgautor()} function. Both
\code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood
which marginalizes over the model's fixed effects and latent mean while
formally incorporating spatial covariance among observations. The inputs
and outputs of \code{spglm()} and \code{spgautor()} closely resemble the
\code{glm()} function from base \proglang{R}, easing the transition from
GLMs to SPGLMs. \pkg{spmodel} provides and builds upon several commonly
used helper functions for model building like \code{summary()},
\code{plot()}, and \code{fitted()}, among others. Spatial predictions of
the latent mean at unobserved locations are obtained using
\code{predict()} or \code{augment()}. \pkg{spmodel} accommodates myriad
advanced modeling features like geometric anisotrpoy, nonspatial random
effects, analysis of variance, and more. Throughout, we use
\pkg{spmodel} to fit SPGLMs to moose presence and counts in Alaska,
United States (US), skewed conductivity data in the Southwestern US,
harbor seal abundance trends in Alaska, US, and voter turnout rates in
Texas, US.
}

\Keywords{autoregressive model, geostatistical model, Poisson
regression, link function, logistic regression, overdispersion, spatial
covariance, spatial dependence}
\Plainkeywords{autoregressive model, geostatistical model, Poisson
regression, link function, logistic regression, overdispersion, spatial
covariance, spatial dependence}

@@

In practice, non-Gaussian data (e.g., binary, count, skewed, and
proportion data) are ubiquitous. Non-Gaussian data that belong to an
exponential family can be naturally modeled using a generalized linear
model (GLM) regression framework
\citep{nelder1972generalized, mccullagh1989generalized}. In a GLM, an
\(n \times 1\) response variable \(\mathbf{y}\) belongs to a statistical
distribution (e.g., binomial, Poisson) with some mean and variance.
Often, the analysis goal is to study the impact of a linear function of
several explanatory variables on the mean of \(\mathbf{y}\) through a
GLM. In this context, the latent (i.e., unobserved) mean of
\(\mathbf{y}\), \(\boldsymbol{\mu}\), is linked to these explanatory
variables via a link function: \begin{align}\label{eq-glm}
f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
\end{align} where for a sample size \(n\), \(f(\cdot)\) is a link
function that connects \(\boldsymbol{\mu}\) to \(\mathbf{w}\),
\(\mathbf{X}\) is the \(n \times p\) design matrix of explanatory
variables, and \(\boldsymbol{\beta}\) is the \(p \times 1\) vector of
fixed effects. While the mean is typically constrained in some way
(e.g., if a probability, between zero and one), the link function
generally makes \(\mathbf{w}\) unconstrained. Common link functions
include the log odds (i.e., logit) link for binary and proportion data
and the log link for count and skewed data. Equation\(~\)\ref{eq-glm}
can also be written in terms of the inverse link function,
\(f^{-1}(\cdot)\): \begin{align}\label{eq-glm2}
\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta} \equiv f^{-1}(\mathbf{w}) = f^{-1}(\mathbf{X} \boldsymbol{\beta}).
\end{align}

The GLM fixed effects (\(\boldsymbol{\beta}\)) are typically estimated
via maximum likelihood \citep{chambers1992S}. It is often convenient to
compute the maximum likelihood estimates using the iteratively
reweighted least squares (IRWLS) algorithm \citep{wood2017generalized},
which is an approach used by the \code{glm()} function in the
\proglang{R} programming language \citep{rcore2024}. GLMs add an
additional layer of complexity compared to linear regression models, as
the left-hand side of Equation\(~\)\ref{eq-glm} is a function of the
mean of \(\mathbf{y}\) rather than \(\mathbf{y}\) itself (as in linear
regression models).

The standard GLM assumes the elements of \(\mathbf{y}\) are independent.
This independence assumption is typically impractical for spatial data.
For spatial data, nearby observations tend to be more similar than
distant observations \citep{tobler1970computer}, which leads to positive
spatial covariance among observations. The consequences of ignoring
spatial covariance in statistical models for spatial data can be severe
and include imprecise parameter estimates as well as misleading standard
errors that inflate Type-I error rates and decrease power
\citep{zimmerman2024spatial}.

An approach for handling spatial data using a GLM is to assume the
elements of \(\mathbf{w}\) exhibit both spatial and nonspatial
variability. This is achieved by adding to Equation\(~\)\ref{eq-glm} two
random effects, \(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\). The
random effect \(\boldsymbol{\tau}\) is an \(n \times 1\) column vector
of spatially dependent random errors. We assume that
\(\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}\) and
\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}\), where
\(\text{E}(\cdot)\) and \(\text{Cov}(\cdot)\) denote expectation and
covariance, respectively. The variance parameter \(\sigma^2_\tau\)
controls the magnitude of spatial covariance and is often called a
partial sill. The matrix \(\mathbf{R}\) is an \(n \times n\) spatial
correlation matrix that depends on a range parameter controlling the
distance-decay rate of the spatial correlation. One example of a spatial
covariance matrix is the ``exponential,'' which is given by
\begin{align}\label{eq-spcov-exp}
  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{\tau} \mathbf{R}_{exp} =  \sigma^2_{\tau} \exp(-\mathbf{H}/\phi),
\end{align} where \(\mathbf{H}\) is a matrix of pairwise distances among
the elements of \(\mathbf{y}\) and \(\phi\) is the range parameter. From
Equation\(~\)\ref{eq-spcov-exp}, as the distance between two elements of
\(\mathbf{y}\) increases, the spatial covariance decreases, which
reflects intuition. Moreover, as the range parameter, \(\phi\),
increases, the strength of spatial dependence increases
(Figure\(~\)\ref{fig-range}). The random effect
\(\boldsymbol{\epsilon}\) is an \(n \times 1\) column vector of
independent random errors. We assume that
\(\text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}\) and
\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\epsilon \mathbf{I}\), where
\(\mathbf{I}\) is an \(n \times n\) identity matrix. The variance
parameter \(\sigma^2_\epsilon\) controls the magnitude of nonspatial
variability (i.e., fine-scale variation) and is often called a nugget.
Often in spatial statistics, quantities are explicitly referenced with
respect to \(\mathbf{s}\), a vector of spatial coordinates indexing the
observation \citep{cressie1993statistics}. For example, \(\mathbf{y}\)
and \(\mathbf{X}\) may instead be written \(\mathbf{y}(\mathbf{s})\) and
\(\mathbf{X}(\mathbf{s})\), respectively. We acknowledge the utility of
this nomenclature but drop the explicit dependence on \(\mathbf{s}\) for
simplicity of notation moving forward.

@@

Fitting and using SPGLMs is challenging both conceptually and
computationally \citep{bolker2009generalized}. Recently, however, there
have been numerous, significant advances in \proglang{R} software that
have made these models more accessible to practitioners. The \pkg{brms}
\citep{burkner2017brms}, \pkg{carBayes} \citep{lee2013carbayes},
\pkg{ngspatial} \citep{hughes2020ngspatial}, \pkg{R-INLA}
\citep{lindgren2015bayesian}, \pkg{inlabru} \citep{bachl2019inlabru},
\pkg{spBayes} \citep{finley2007spbayes}, \pkg{spOccupancy}
\citep{doser2022spoccupancy}, \pkg{spAbundance}
\citep{doser2024spabundance}, and \pkg{spNNGP} \citep{finley2002spnngp}
packages take a Bayesian approach, either directly sampling from
posterior distributions of parameters (e.g., using MCMC) or
approximating them. A benefit of Bayesian approaches is that prior
information can be incorporated and uncertainty quantification of
parameter estimates is straightforward. However, Bayesian approaches,
especially those using MCMC, can be computationally expensive. In order
to reduce computation time, many of these packages (e.g., \pkg{R-INLA})
work with the precision matrix instead of the covariance matrix so that
computationally expensive matrix inversion is not required. Working with
precision matrices, however, can be more restrictive and less intuitive
than working with covariance matrices. The \pkg{FRK}
\citep{sainsbury2024modeling}, \pkg{glmmTMB} \citep{brooks2017glmmtmb},
\pkg{hglm} \citep{ronnegard2010hglm}, \pkg{mgcv}
\citep{wood2017generalized}, and \pkg{spaMM} \citep{rousset2014spamm}
packages directly use Laplace, quasi-likelihood, or reduced-rank
approaches to estimate parameters. These direct approaches tend to be
computationally efficient, as they don't rely on MCMC sampling. In
contrast to the Bayesian approach, a drawback of these direct approaches
is that prior information cannot be formally incorporated and covariance
parameter uncertainty is more challenging to quantify. The \pkg{sdmTMB}
\citep{anderson2024sdmtmb} package combines elements of \pkg{R-INLA},
\pkg{glmmTMB}, and Gaussian Markov random fields to fit a wide variety
of SPGLMs, while \pkg{tinyVAST} \citep{thorson2025tinyVAST} extends some
of these models to multivariate or (dynamic) structural equation models.

Building from \citet{evangelou2011estimation} and
\citet{bonat2016practical}, \citet{ver2024marginal} proposed a novel
approach for fitting SPGLMs that leverages the Laplace approximation
while marginalizing over both the latent \(\mathbf{w}\) and the fixed
effects (\(\boldsymbol{\beta}\)). This approach performed well in a
variety of simulation settings, generally having appropriate confidence
interval coverage for the fixed effects and prediction interval coverage
for \(\mathbf{w}\) at new locations. It also performed similarly to the
Bayesian SPGLM approach in \pkg{spBayes} and the automatic
differentiation SPGLM approach in \pkg{glmmTMB} but was much faster. At
small sample sizes, the approach outperformed the approximate Bayesian
SPGLM approach in \pkg{R-INLA} and had similar computational times. For
moderate sample sizes, it performed similarly to \pkg{R-INLA}, though
\pkg{R-INLA} was faster. The novel Laplace approach is particularly
attractive for two reasons. First, it is general enough that it can be
applied to any covariance structure (not just spatial). Second, after
estimating the covariance parameters, analytical solutions exist for the
fixed effects (and their standard errors) as well as predictions of the
latent \(\mathbf{w}\) at new locations (and their standard errors).

The \pkg{spmodel} \proglang{R} package \citep{dumelle2023spmodel}
recently released a full set of modeling tools for SPGLMs fit using the
novel Laplace approach described by \citet{ver2024marginal}. These
modeling tools are approachable and mirror the familiar \code{glm()}
syntax from base-\proglang{R}, making the transition from GLMs to SPGLMs
relatively seamless. The \code{spglm()} function fits SPGLMs for
point-referenced support (e.g., x- and y-coordinates representing point
locations in a field; these models are sometimes called
``geostatistical'' models), while the \code{spgautor()} function fits
SPGLMs for areal support (e.g., polygon boundaries representing
geographic subsets of a region; these models are sometimes called
``autoregressive'' models). For both point-referenced and areal
supports, \pkg{spmodel} supports the binomial distribution for binary
data, Poisson and negative binomial distributions for count data, Gamma
and inverse Gaussian distributions for skewed data, and the beta
distribution for proportion data. There are 20 different spatial
covariance structures available including the exponential, Gaussian, and
spherical for point-referenced support (Figure\(~\)\ref{fig-type}) and
the conditional autoregressive, and simultaneous autoregressive
structures for areal support \pkg{spmodel} provides tools for commonly
used model summaries, visualizations, and diagnostics (e.g., fitted
values) using standard \proglang{R} helper functions like
\code{summary()}, \code{plot()}, and \code{fitted()}, among others.
\pkg{spmodel} also provides tools to predict \(\mathbf{w}\) at new
locations and quantify uncertainty in those prediction using
\code{predict()} and \code{augment()}. This core functionality, combined
with several advanced features we describe throughout the manuscript,
enables \pkg{spmodel} to introduce novel, important SPGLM modeling tools
previously missing from the existing \proglang{R} ecosystem.

@@

Of the existing \proglang{R} packages for SPGLMs, \pkg{spmodel} (version
0.11.1) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in
terms of scope and feel. Both packages use similar syntax as
\code{glm()}, accommodate flexible \code{formula} arguments (e.g.,
offsets, splines), handle spatial covariance that decays at different
rates in different directions (i.e., geometric anisotropy), incorporate
nonspatial random effects, support other \proglang{R} packages for
modeling like \pkg{broom} \citep{robinson2021broom, kuhn2022tidy},
\pkg{emmeans} \citep{lenth2024emmeans}, and \pkg{car}
\citep{fox2019car}, and have tools for model summaries, prediction, and
simulating data. There are some notable differences between the two
packages, however. \pkg{sdmTMB} supports several additional GLM
distributions like the Tweedie, supports Hurdle models, and can
incorporate prior information through Bayesian applications.
\pkg{sdmTMB} also provides tools for working with temporal data and
spatiotemporal data and provides enhanced visualizations of the model's
marginal effects. \pkg{sdmTMB} does require a preprocessing step of
constructing a mesh prior to model fitting (using the stochastic partial
differential equation approach), and the density of the mesh can affect
model results and computational complexity. On the other hand,
\pkg{spmodel} does not require the construction of a mesh prior to model
fitting. \pkg{spmodel} supports 20 different spatial covariances and
models them directly, rather than using a precision matrix approximation
to the Matérn spatial covariance as in \pkg{sdmTMB}. \pkg{spmodel} can
model data directly using neighborhood distance and autoregressive
models, rather than relying on the polygon centroid (as in
\pkg{sdmTMB}), which may not be within the polygon's boundaries.
\pkg{spmodel} provides experimental design tools (e.g., analysis of
variance, contrasts), supports \pkg{sf} objects in modeling and
prediction functions \citep{pebesma2018sf}, has several specialized
model diagnostics like leverage values and Cook's distances, and has
analytic solutions for fixed effect and prediction standard errors.
Other similarities and differences do exist between \pkg{sdmTMB} and
\pkg{spmodel}, and both packages continue to evolve. Overall, we believe
that these packages are complementary and enhance the suite of SPGLM
tools accessible to practitioners.

The rest of this article is organized as follows. In
Section\(~\)\ref{sec-spglm}, we provide some background for the SPGLM
fitting and prediction routines in \pkg{spmodel}. In
Section\(~\)\ref{sec-applications}, we provide an overview of core SPGLM
functionality in \pkg{spmodel} by modeling moose presence in Alaska,
United States (US). In Section\(~\)\ref{sec-applications2}, we model
moose counts in Alaska, US; skewed lake conductivity in the Southwestern
US; harbor seal abundance trend behavior in Alaska, US; and voter
turnout rates in Texas, US. And in Section\(~\)\ref{sec-discussion}, we
end with a discussion synthesizing \pkg{spmodel}'s contributions to
SPGLMs in \proglang{R}.

@@

The novel Laplace approach implemented in \pkg{spmodel} formally
maximizes a hierarchical GLM likelihood
\citep{lee1996hierarchical, wood2017generalized}, making
likelihood-based statistics for model comparison like AIC
\citep{akaike1974new}, AICc \citep{hoeting2006model}, BIC
\citep{schwarz1978estimating}, deviance
\citep{mccullagh1989generalized}, and likelihood ratio tests available.
These types of statistics are not available for quasi-likelihood
\citep{wedderburn1974quasi, breslow1993approximate} or pseudo-likelihood
approaches \citep{wolfinger1993generalized}, which only specify the
first two moments of a distribution. Next, we describe a brief overview
of the approach and how it can be used for several primary data analysis
tasks \citep{tredennick2021practical} like model comparison, parameter
estimation, inference, model diagnostics, and prediction.

@@

  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{w}^{*}), \varphi] [\mathbf{w}^{*} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{w}^{*}}|^{-1/2},
\end{align} which can be directly evaluated. This result suggests a
doubly iterative optimization over 1) \(\boldsymbol{\theta}\) and
\(\varphi\) and 2) the latent \(\mathbf{w}\) (to find each set of
\(\mathbf{w}^{*}\)), which ultimately yields the marginal restricted
maximum likelihood estimators \(\hat{\varphi}\) and
\(\hat{\boldsymbol{\theta}}\) and their respective values of
\(\mathbf{w}^{*}\), which we call \(\hat{\mathbf{w}}\).
\citet{ver2024marginal} provide further details, which includes explicit
forms of \(\mathbf{g}\) and \(\mathbf{G}\) for various response
distributions.

\subsection{Estimating fixed effects}\label{sec:fixed}
